Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2024-06-14 plos_latex_template.tex
add paragraph state variable

One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. For instance, if we model a process as Poisson, we assume that if the mean of the process is large, the variance will be large as well. Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings. Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accomodated by a Poisson model. The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model. Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).

@@

\begin{eqnarray}
\label{eq:schemeP}
	\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)
\end{eqnarray}

@@

\begin{eqnarray}
\label{eq:schemeP}
    log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)

    log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)

    log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)
\end{eqnarray}

So, the form of the probability mass function is:
$$f_t(I) = {I + \theta - 1 \choose I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta$$
This form has expectation and variance as follows:
$$E(I) = \mu$$
$$Var(I) = \mu + \frac{\mu^2}{\theta}$$

@@

\begin{figure}[!h]
\caption{{\bf Fig 2.}
Incidence and dispersion between 2020-02-20 and 2023-03-19 in large counties in the US. A: Binned log of the dispersion parameter over time. B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis). C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis). D: LRT p-values over time as well as for each of the large counties (y-axis).
}
\label{fig1}
\end{figure}

The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variability than when incidence is lower. In other words, what makes a change in dispersion meaningful is how it affects variance and variance-mean relationships. For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. A change from $\theta = 1000$ to $\theta = 100$ is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
$$var = \mu + \mu^2/\theta$$
So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing (Dalziel et al. 2018), or increased locally experienced mean density (Lloyd 1967). To illustrate this using the empirical data, we observed an increase from $\theta$ = 2.707735e+16 to $\theta = 21.8711$, which, conditional on the observed case count of 3$$, would lead to an increase in variance from around 3 to around 3.4. In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.

\begin{figure}[!h]
\caption{{\bf Fig 3.}

}
\label{fig1}
\end{figure}

>>>>

One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. For instance, if we model a process as Poisson, we assume that if the mean of the process is large, the variance will be large as well. Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings. Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accomodated by a Poisson model. The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model. In distributions that form part of the exponential family, the dispersion parameter determines the mean-variance relationship.

$$f(y_i; \theta_i, \phi) = exp\{\theta_i y_i - b(\theta_i)/a_i(\phi) - c(y_i; \phi)\}, i = 1,...,n$$
$$Var(y_i) = b''(\theta_i) a_i(\phi) $$

Where $\theta_i$ is the natural (canonical) parameter and $\phi$ is the dispersion parameter.

Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).

@@

    $$\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)$$

@@

    $$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
    $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
    $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$

So, the form of the probability mass function is:

$$f_t(I) = {I + \theta - 1 \choose I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta$$

This form has expectation and variance as follows:

$$E(I) = \mu$$
$$Var(I) = \mu + \frac{\mu^2}{\theta}$$

@@

\begin{figure}[!h]
\caption{{\bf Fig 2.}
Incidence and dispersion between 2020-02-20 and 2023-03-19 in large counties in the US. A: Binned log of the dispersion parameter over time. B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis). C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis). D: LRT p-values over time as well as for each of the large counties (y-axis).
}
\label{fig2}
\end{figure}

The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variability than when incidence is lower. In other words, what makes a change in dispersion meaningful is how it affects variance and variance-mean relationships. For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. A change from $\theta = 1000$ to $\theta = 100$ is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:

$$var = \mu + \mu^2/\theta$$

So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing (Dalziel et al. 2018), or increased locally experienced mean density (Lloyd 1967). To illustrate this using the empirical data, we observed an increase from $\theta = 2.707735e+16$ to $\theta = 21.8711$, which, conditional on the observed case count of 3, would lead to an increase in variance from around 3 to around 3.4. In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.

Additionally, $\theta$ seems to behave like a state variable. That is, its rate of change is dependent on its value and the values of other state variables (like incidence), which determine the "state" of the system. State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics. This predictability and simplicity are crucial for both theoretical studies and practical applications, and creating a reliable description of this (epidemic) system is overdue.

\begin{figure}[!h]
\caption{{\bf Fig 3.}
caption
}
\label{fig3}
\end{figure}
